---
title: "Evaluate projections"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Evaluate projections}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(propop)
```
# Evaluate projections

`propop` provides a function to **compare projections against a benchmark**. 
The benchmark can be the actual population development (i.e., official records) 
or a base model (e.g., when testing alternative models). 

## Get the data

To run the evaluation function, we need the benchmark data and the 
results from the projection.  

Get the *population records* for the canton of Aargau 2019-2022:  

``` {r get-benchmark, message = FALSE}
data_benchmark <- get_population(
  number_fso = "px-x-0102010000_101",
  year_first = 2018,
  year_last = 2022,
  spatial_units = "- Aargau"
)
```

Get the parameters for the projection:

``` {r get-parameters, message = FALSE, warning=FALSE, results='hide'}
data_parameters <- get_parameters(
  year_first = 2019,
  year_last = 2050,
  spatial_units = "Aargau"
)
```

Get the *population projection* for 2019-2022:

``` {r project-canton}
data_projected <- propop(
  parameters = data_parameters |>
    dplyr::filter(scen == "reference"),
  year_first = 2019,
  year_last = 2050,
  age_groups = 101,
  fert_first = 16,
  fert_last = 50,
  share_born_female = 100 / 205,
  # population records from 2018 as starting point
  population = data_benchmark |>
    dplyr::filter(year == 2018),
  subregional = FALSE
)
```

## Evaluate 1-year age classes

Let's first pit the projected population growth against the recorded population
development (= benchmark) using one-year age classes (default option). Make sure
to pass matching data frames, especially in terms of the year ranges. To remove
the start population (2018) from the projection results, you could also use
`drop_start_year = TRUE`. 

As a result of `prepare_evaluation()`, you get the recorded (`n_bench`) and 
projected (`n_proj`) population for each demographic group and year next to each
other: 

``` {r combine}
# Combine and pre-process the data
combined <- prepare_evaluation(
  # only keep years from projected period
  data_benchmark = data_benchmark |> dplyr::filter(year > 2018),
  data_projected = data_projected |> dplyr::filter(year > 2018 & year <= 2022)
)

# Show combined data
combined
```

Based on the difference between the observed and projected number of people, 
`compute_measures()` then calculates the error and several performance metrics:

``` {r evaluate1}
compute_measures(combined) |>
  # select demographic group
  dplyr::filter(sex == "m" & nat == "ch" & age == 27) |>
  # round to two digits
  dplyr::mutate(across(perc_e:abs_perc_e, \(x) sprintf(fmt = "%.2f", x))) |>
  DT::datatable(
    filter = "top",
    caption = htmltools::tags$caption(
      style = "caption-side: top; text-align: left; font-weight: bold",
      "Performance measures for projections with 1-year age classes for the
      canton of Aargau 2019-2022."
    )
  )
```

## Evaluate larger age groups

You may also conduct the evaluation for the commonly used age groups 
0-19 year olds, 20-64 year olds, and over 64 year olds using the option 
`age_groups = "age_groups_3"`.

``` {r combine-grouped}
# Combine and pre-process the data
combined_grouped <- prepare_evaluation(
  # only keep years from projected period
  data_benchmark = data_benchmark |> dplyr::filter(year > 2018),
  data_projected = data_projected |> dplyr::filter(year > 2018 & year <= 2022),
  age_groups = "age_groups_3"
)

# Show combined data
combined_grouped

# Compute the performance measures
compute_measures(combined_grouped) |>
  # select demographic group
  dplyr::filter(sex == "m" & nat == "ch" & age == "age_20_64") |>
  # round to two digits
  dplyr::mutate(across(perc_e:abs_perc_e, \(x) sprintf(fmt = "%.2f", x))) |>
  DT::datatable(
    filter = "top",
    caption = htmltools::tags$caption(
      style = "caption-side: top; text-align: left; font-weight: bold",
      "Performance measures for projections with three age classes for the
      canton of Aargau 2019-2022."
    )
  )
```

## Using group weights

Furthermore, you can use weights. The adjusted metric `w_abs_perc_e` considers 
that smaller groups tend to have larger forecast errors. As a result, adjusted
absolute percentage errors weigh less heavily in smaller groups than in larger 
groups.

``` {r weighted}
compute_measures(
  combined_grouped,
  weight_groups = c("age")
) |>
  dplyr::filter(year == 2019 & abs_perc_e > .34 & abs_perc_e < .39) |>
  dplyr::mutate(across(perc_e:w_abs_perc_e, \(x) sprintf(fmt = "%.2f", x))) |>
  DT::datatable(
    filter = "top",
    caption = htmltools::tags$caption(
      style = "caption-side: top; text-align: left; font-weight: bold",
      "Performance measures for projections with three age classes for the
      canton of Aargau 2019-2022 (incl. weighted measure)."
    )
  )
```

## `propop` versus FSO

Instead of comparing the propop projection with population records, we can also
use the evaluation function to check how close the `propop` projection comes to
FSO's projection from the 2020 model.

The differences are extremely small (>= 1):

``` {r fso-comparison}
fso_propop <- prepare_evaluation(
  data_benchmark = data_parameters |>
    dplyr::filter(scen == "reference" & year > 2018) |>
    dplyr::mutate(n = n_projected),
  data_projected = data_projected |>
    dplyr::filter(year > 2018),
  age_groups = "age_groups_3"
) |>
  compute_measures()

fso_propop |>
  dplyr::mutate(across(perc_e:abs_perc_e, \(x) sprintf(fmt = "%.2f", x))) |>
  DT::datatable(
    filter = "top",
    caption = htmltools::tags$caption(
      style = "caption-side: top; text-align: left; font-weight: bold",
      "Comparison of projetions from FSO versus `propop`."
    )
  )
```

```{r fig.width = 9, fig.height = 7, echo = FALSE}
library(ggplot2)
ggplot(fso_propop) +
  geom_jitter(aes(x = year, y = perc_e), size = 0.7) +
  labs(
    title = "Differences between FSO and propop projections for each demographic group",
    x = "Year",
    y = "Difference in percent"
  ) +
  theme_minimal()
```


