---
title: "Prepare data"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Prepare data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 84
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

This vignette shows how to obtain and prepare the relevant information to run
population projections with the package `propop` at the level of a Canton.

Most of the information and data required to run `propop::propop()` (or
`propop::project_raw()`) can be downloaded from
[STAT-TAB](https://www.pxweb.bfs.admin.ch/pxweb). More specifically, 
the information from the following tables are needed:

+-------------------+--------------------------------+----------------------------+
| Table ID          | Parameters expressed as...     | Variables required for     |
|                   |                                | projection                 |
+===================+================================+============================+
| px-               | number of people (*reference*  | -   Inter-cantonal         |
| x-0104020000_101  | scenario)                      |     immigration            |
|                   |                                | -   Inter-cantonal         |
|                   |                                |     emigration             |
|                   |                                | -   International          |
|                   |                                |     immigration            |
|                   |                                | -   International          |
|                   |                                |     emigration             |
|                   |                                | -   end of year population |
|                   |                                |     size                   |
+-------------------+--------------------------------+----------------------------+
| px-               | number of people (*high        | -   same as 101            |
| x-0104020000_102  | growth* scenario)              |                            |
+-------------------+--------------------------------+----------------------------+
| px-               | number of people (*low growth* | -   same as 101            |
| x-0104020000_103  | scenario)                      |                            |
+-------------------+--------------------------------+----------------------------+
| px-               | rates / probabilities (five    | -   Births per mother      |
| x-0104020000_109  | scenarios)                     |                            |
|                   |                                | -   Mortality              |
|                   |                                |                            |
|                   |                                | -   International          |
|                   |                                |     emigration             |
|                   |                                |                            |
|                   |                                | -   Inter-cantonal         |
|                   |                                |     emigration             |
|                   |                                |                            |
|                   |                                | -   Acquisition of Swiss   |
|                   |                                |     citizenship            |
+-------------------+--------------------------------+----------------------------+
| px                | share of newborns with Swiss   | -   Live newborns          |
| -x-0104020000_106 | nationality born to non-Swiss  |                            |
|                   | mothers                        | -   Live births by age and |
|                   |                                |     nationality of the     |
|                   |                                |     mother (varies between |
|                   |                                |     cantons)               |
+-------------------+--------------------------------+----------------------------+
| Constant          |                                | -   Start (16) and end (50)|
| parameters        |                                |     of the fertile age of  |
| **not** directly  |                                |     women                  |
| available from    |                                |                            |
| STAT-TAB must be  |                                | -   Proportion of newborns |
| provided as       |                                |     with female sex        |
| arguments         |                                |     (100/205)              |
+-------------------+--------------------------------+----------------------------+

: Overview of required FSO tables (STAT-TAB)


# Convenient way to get FSO data

The `propop` package provides two convenience functions to download data from 
the FSO. 

To get the starting population for a spatial unit, you must use the spelling
defined in the corresponding FSO table. The entries in the FSO tables may 
contain special characters. The spelling may also vary between FSO tables. 

`BFS::bfs_get_metadata()` is helpful to identify the required spelling(s) 
(see further down on this page).

Here's an example of how to get the population for the canton of Aargau:
```{r eval = FALSE}
library(propop) 
ag_population <- get_population(
  number_fso = "px-x-0102010000_101",
  year_first = 2022,
  year_last = 2022,
  spatial_units = "- Aargau"
)
```

Get the parameters for a sample canton (mind using the same spelling as in the
FSO tables; see comment above):
```{r get ag params, eval = FALSE}
ag_parameters <- get_parameters(
  year_first = 2023,
  year_last = 2026,
  spatial_units = c("Aargau")
)
```
The projection can be run as follows:

```{r eval = FALSE}
# select reference scenario
ag_parameters_ref <- ag_parameters |>
  dplyr::filter(scen == "reference")

propop(
  parameters = ag_parameters_ref,
  year_first = 2023,
  year_last = 2026,
  age_groups = 101,
  fert_first = 16,
  fert_last = 50,
  share_born_female = 100 / 205,
  population = ag_population,
  subregional = FALSE,
  binational = TRUE
)
```


**Note of caution:** 
As long as the FSO's API interface and the underlying data structure remain 
stable, the functions will work. However, changes in the API are likely to break 
the functions.
             

# Manual way to get FSO data

In case the above shouldn't work or if you want to retrace the necessary steps
manually, we also provide a step-by-step description of how to get the 
population data and the projection parameters from the FSO.
             
# Packages

To download the data, we need the following packages:

```{r load packages, , echo= T, eval=T, results='hide', message=FALSE, warning=FALSE}
library(BFS) # to download FSO data
library(dplyr) # to process data
library(tidyr) # to wrangle data
library(propop) # to run population projection
```
                    
                                
# Prepare queries and download data
                                
To make the data download faster, save disk space, and avoid filtering after the
download, it is advisable to specify and download only the information that we
really need. To prepare such a customised, reduced data download, the
[instructions](https://felixluginbuhl.com/BFS/#query-specific-elements) from the
`BFS` package are very helpful.
                                
Following these instructions, we can use the `text` and `valueTexts` variables 
to generate a **query** dimension object for each table and to **download** 
the data (see following subsections). To illustrate, for table 
`px-x-0104020000_101`, we can obtain the meta data as follows:
                                
```{r get meta data, echo= T, eval=TRUE,  message=FALSE}
metadata <- BFS::bfs_get_metadata(number_bfs = "px-x-0104020000_101")
metadata_tidy <- metadata |>
  select(-valueTexts) |>
  unnest_longer(values) |>
  dplyr::mutate(
    valueTexts = metadata |>
      select(valueTexts) |>
      unnest_longer(valueTexts) |>
      pull(valueTexts)
  ) |>
  select(code, text, values, valueTexts, everything())

head(metadata_tidy)
```
                                
Although the structure of the first three tables should be identical, the low growth
scenario (`_103`) contains different meta information and requires some changes.
                                
## Get "number of people" parameters
                                
Some of FSO's expectations are expressed in "number of people" parameters (first
three entries in the table). These parameters indicate FSO expectations about 
how many people do certain things (e.g., how many 64-year old Swiss men will 
emigrate to another country in 2043).

To prepare the download of these parameters, we can specify the following query:

```{r filter dimensions, echo = T, eval = FALSE, results='hide', message=FALSE, warning=FALSE}
# Specify the elements to download
dim1 <- metadata_tidy |>
  dplyr::filter(
    text == "Kanton" & # Canton
      valueTexts %in% c("Aargau")
  )

dim2 <- metadata_tidy |>
  dplyr::filter(
    text == "Geschlecht" & # sex
      valueTexts %in% c(
        "Mann", # male
        "Frau"
      )
  ) # female

dim3 <- metadata_tidy |>
  dplyr::filter(
    text == "Alter" & # get each age group
      !(valueTexts %in% "Alter - Total")
  ) # but exclude "Total"

dim4 <- metadata_tidy |>
  dplyr::filter(text == "Jahr") # get all years

# adapt to the different structure of the "low" scenario table
dim4_103 <- metadata_tidy |>
  dplyr::filter(
    text == "Jahr"
  ) |> # get all years
  dplyr::mutate(values = as.character(0:31))

dim5 <- metadata_tidy |>
  dplyr::filter(
    text == "Staatsangehörigkeit (Kategorie)" & # nationality
      valueTexts %in% c(
        "Schweiz", # Swiss
        "Ausland"
      )
  ) # Foreign / international

dim6 <- metadata_tidy |>
  dplyr::filter(
    text == "Beobachtungseinheit" & # parameters for projection
      valueTexts %in% c(
        "Einwanderungen", # international immigration
        "Auswanderungen", # international emigration
        "Interkantonale Zuwanderungen", # inter-cantonal immigration
        "Interkantonale Abwanderungen" # inter-cantonal emigration
      )
  )

# build dimensions list object
dimensions <- list(
  dim1$values,
  dim2$values,
  dim3$values,
  dim4$values,
  dim5$values,
  dim6$values
)
# add names
names(dimensions) <- c(
  unique(dim1$code),
  unique(dim2$code),
  unique(dim3$code),
  unique(dim4$code),
  unique(dim5$code),
  unique(dim6$code)
)

# version for _103
# build dimensions list object
dimensions_103 <- list(
  dim1$values,
  dim2$values,
  dim3$values,
  dim4_103$values,
  dim5$values,
  dim6$values
)

# add names
names(dimensions_103) <- c(
  unique(dim1$code),
  unique(dim2$code),
  unique(dim3$code),
  unique(dim4_103$code),
  unique(dim5$code),
  unique(dim6$code)
)
```

Using the above specifications, we can **download** the FSO "number of people"
parameters as follows:

```{r query, eval = FALSE, echo = TRUE, message=FALSE}
# reference scenario
fso_numbers_r <- BFS::bfs_get_data(
  number_bfs = "px-x-0104020000_101",
  query = dimensions
) |>
  rename(value = paste0(
    "Szenarien zur Bevölkerungsentwicklung der Kantone 2020-2050,",
    " Referenzszenario AR-00-2020 - zukünftige Bevölkerungsentwicklung"
  )) |>
  dplyr::mutate(scen = "reference")

# high growth scenario
fso_numbers_h <- BFS::bfs_get_data(
  number_bfs = "px-x-0104020000_102",
  query = dimensions
) |>
  rename(value = paste0(
    "Szenarien zur Bevölkerungsentwicklung der Kantone 2020-2050,",
    " 'hohes' Szenario BR-00-2020 - zukünftige Bevölkerungsentwicklung"
  )) |>
  dplyr::mutate(scen = "high")

# low growth scenario
fso_numbers_l <- BFS::bfs_get_data(
  number_bfs = "px-x-0104020000_103",
  query = dimensions_103
) |>
  rename(value = paste0(
    "Szenarien zur Bevölkerungsentwicklung der Kantone 2020-2050,",
    " 'tiefes' Szenario CR-00-2020 - zukünftige Bevölkerungsentwicklung"
  )) |>
  dplyr::mutate(scen = "low")

# combine into a single data frame
fso_numbers_raw <- full_join(fso_numbers_r, fso_numbers_h) |>
  full_join(fso_numbers_l)
```

```{r process, eval = FALSE, echo = FALSE}
# Bring variable names and factor levels into the format required later
fso_numbers <- fso_numbers_raw |>
  dplyr::rename(
    nat = "Staatsangehörigkeit (Kategorie)",
    sex = Geschlecht,
    age = Alter,
    year = Jahr,
    fso_parameter = Beobachtungseinheit
  ) |>
  # change factor levels
  dplyr::mutate(
    fso_parameter = case_match(
      fso_parameter,
      "Auswanderungen" ~ "emi_n",
      "Einwanderungen" ~ "imm_int",
      "Interkantonale Abwanderungen" ~ "interc_emi",
      "Interkantonale Zuwanderungen" ~ "interc_imm"
    ),
    nat = case_match(
      nat,
      "Schweiz" ~ "ch",
      "Ausland" ~ "int"
    ),
    sex = case_when(
      sex == "Mann" ~ "m",
      sex == "Frau" ~ "f"
    ),
    age = as.numeric(stringr::str_extract(age, "\\d+"))
  )
```

```{r quick-check-numbers, echo = F, eval=FALSE}
# quick check to see if figures match STAAG input table for projection
fso_numbers |>
  dplyr::filter(
    year == 2029 &
      scen == "reference" &
      nat == "int" &
      sex == "f" &
      age == 0
  )
```

## Get "rates" and "probabilities"

The FSO indicates some of its expectations as "rates" or "probabilities" (row four
in the overview table at the top). To illustrate, these parameters could indicate
the likelihood of 24-year old Swiss women to have a child in the year 2034.

Before we can download the data, we again need the metadata:

```{r get meta data2, echo= T, eval=TRUE,  message=FALSE, warning=FALSE}
metadata <- BFS::bfs_get_metadata(number_bfs = "px-x-0104020000_109")

metadata_tidy <- metadata |>
  select(-valueTexts) |>
  unnest_longer(values) |>
  dplyr::mutate(
    valueTexts = metadata |>
      select(valueTexts) |>
      unnest_longer(valueTexts) |>
      pull(valueTexts)
  ) |>
  select(code, text, values, valueTexts, everything())

head(metadata_tidy)
```

To **download** the "rate" and "probability" parameters (last row in the table), we
can use the following specifications:

```{r filter dimensions2, echo = T, eval = FALSE, results='hide', message=FALSE, warning=FALSE}
# Specify the elements to download
dim1 <- metadata_tidy |>
  dplyr::filter(
    text == "Kanton" & # Canton
      valueTexts %in% c("Aargau")
  )

dim2 <- metadata_tidy |>
  dplyr::filter(
    text == "Szenario-Variante" & # sex
      valueTexts %in% c(
        "Referenzszenario AR-00-2020", # reference scenario
        "'hohes' Szenario BR-00-2020", # high growth
        "'tiefes' Szenario CR-00-2020"
      )
  ) # low growth

dim3 <- metadata_tidy |>
  dplyr::filter(
    text == "Staatsangehörigkeit (Kategorie)" & # nationality
      valueTexts %in% c(
        "Schweiz", # Swiss
        "Ausland"
      )
  ) # Foreign / international

dim4 <- metadata_tidy |>
  dplyr::filter(
    text == "Geschlecht" & # sex
      valueTexts %in% c(
        "Mann", # male
        "Frau"
      )
  ) # female

dim5 <- metadata_tidy |>
  dplyr::filter(
    text == "Alter" & # all 1-year age groups
      !(valueTexts %in% "Alter - Total")
  ) # but exclude "Total"

dim6 <- metadata_tidy |>
  dplyr::filter(
    text == "Jahr"
  ) # get all years

dim7 <- metadata_tidy |>
  dplyr::filter(
    text == "Beobachtungseinheit" & # type of parameter types
      valueTexts %in% c(
        "Geburtenziffern", # births
        "Prospektive Sterbewahrscheinlichkeiten", # mortality
        "Auswanderungsziffern", # international emigration
        "Interkantonale Abwanderungsziffern", # inter-cantonal emigration
        "Einbürgerungsziffern"
      )
  ) # acquisition of Swiss citizenship

# build dimensions list object
dimensions <- list(
  dim1$values,
  dim2$values,
  dim3$values,
  dim4$values,
  dim5$values,
  dim6$values,
  dim7$values
)
# add names
names(dimensions) <- c(
  unique(dim1$code),
  unique(dim2$code),
  unique(dim3$code),
  unique(dim4$code),
  unique(dim5$code),
  unique(dim6$code),
  unique(dim7$code)
)
```

Using the above specifications, we can download the FSO "rate" parameters as
follows:

```{r query2, eval = FALSE, message=FALSE,results='hide'}
# Download rate parameters
fso_rates_raw <- BFS::bfs_get_data(
  number_bfs = "px-x-0104020000_109",
  query = dimensions
)
```

We need to process the data to ensure that the structure of the rate parameters
conforms to the expectations of the projection function:

```{r process query2, eval = FALSE, message=FALSE}
# Bring variable names and factor levels into the format required later
fso_rates <- fso_rates_raw |>
  dplyr::rename(
    nat = "Staatsangehörigkeit (Kategorie)",
    sex = Geschlecht,
    age = Alter,
    year = Jahr,
    fso_parameter = Beobachtungseinheit,
    scen = "Szenario-Variante",
    value =
      "Szenarien zur Bevölkerungsentwicklung der Kantone 2020-2050 - Ziffern"
  ) |>
  # change factor levels
  dplyr::mutate(
    scen = case_match(
      scen,
      "Referenzszenario AR-00-2020" ~ "reference",
      "'hohes' Szenario BR-00-2020" ~ "high",
      "'tiefes' Szenario CR-00-2020" ~ "low"
    ),
    nat = case_match(
      nat,
      "Schweiz" ~ "ch",
      "Ausland" ~ "int"
    ),
    sex = case_when(
      sex == "Mann" ~ "m",
      sex == "Frau" ~ "f"
    ),
    age = as.numeric(stringr::str_extract(age, "\\d+")),
    fso_parameter = case_match(
      fso_parameter,
      "Prospektive Sterbewahrscheinlichkeiten" ~ "mor",
      "Auswanderungsziffern" ~ "emi",
      "Interkantonale Abwanderungsziffern" ~ "intercant",
      "Einbürgerungsziffern" ~ "acq",
      "Geburtenziffern" ~ "birth_rate"
    )
  )
```


## Get share of newborns with Swiss nationality born to non-Swiss mothers

Mothers who do not have Swiss citizenship may have a Swiss partner. Their 
children will be Swiss nationals. To use this information in the projection, 
we need to download another table and compute the corresponding parameter.

Before we can download the data, we again need the metadata:

```{r get meta data3, echo= T, eval=FALSE,  message=FALSE, warning=FALSE}
# Get meta data to determine what to download
metadata <- BFS::bfs_get_metadata(number_bfs = "px-x-0104020000_106")
metadata_tidy <- metadata |>
  dplyr::select(-valueTexts) |>
  tidyr::unnest_longer(values) |>
  dplyr::mutate(
    valueTexts = metadata |>
      dplyr::select(valueTexts) |>
      tidyr::unnest_longer(valueTexts) |>
      dplyr::pull(valueTexts)
  ) |>
  dplyr::select(code, text, values, valueTexts, everything())
```

To **download** the "rate" and "probability" parameters (last row in the table), 
we can use the following specifications:

```{r filter dimensions3, echo = T, eval = FALSE, results='hide', message=FALSE, warning=FALSE}
# Specify the elements to download
dim1 <- metadata_tidy |>
  dplyr::filter(
    text == "Kanton" & # Canton
      valueTexts %in% c("Aargau")
  )

dim2 <- metadata_tidy |>
  dplyr::filter(
    text == "Szenario-Variante" & # scenario
      valueTexts %in% c(
        "Referenzszenario AR-00-2020",
        "'hohes' Szenario BR-00-2020",
        "'tiefes' Szenario CR-00-2020"
      )
  )
dim3 <- metadata_tidy |>
  dplyr::filter(
    text == "Staatsangehörigkeit (Kategorie)" & # nationality
      valueTexts %in% "Ausland"
  )
dim4 <- metadata_tidy |>
  dplyr::filter(
    text == "Geschlecht" & # sex
      valueTexts %in% "Geschlecht - Total"
  )
dim5 <- metadata_tidy |>
  dplyr::filter(
    text == "Altersklasse" & # age
      valueTexts %in% "Altersklasse - Total"
  )
dim6 <- metadata_tidy |>
  dplyr::filter(
    text == "Jahr"
  ) # get all years
dim7 <- metadata_tidy |>
  dplyr::filter(
    text == "Beobachtungseinheit" & # type of parameter types
      valueTexts %in%
        c(
          # live births of international newborns to international mothers
          "Lebendgeburten",
          "Lebendgeburten nach Alter und Staatsangehörigkeit der Mutter"
        )
  ) # all live newborns to international mothers

# build dimensions list object
dimensions <- list(
  dim1$values,
  dim2$values,
  dim3$values,
  dim4$values,
  dim5$values,
  dim6$values,
  dim7$values
)

# add names
names(dimensions) <- c(
  unique(dim1$code),
  unique(dim2$code),
  unique(dim3$code),
  unique(dim4$code),
  unique(dim5$code),
  unique(dim6$code),
  unique(dim7$code)
)
```

Using the above specifications, we can download the parameter as
follows:

```{r query3, eval = FALSE, message=FALSE,results='hide'}
# Download rate parameters
fso_births_int_ch_raw <- BFS::bfs_get_data(
  number_bfs = "px-x-0104020000_106",
  query = dimensions
)
```

We need to process the data to ensure that the structure of the parameter
conforms to the expectations of the projection function:

```{r process query3, eval = FALSE, message=FALSE}
# Bring variable names and factor levels into the format required later
# Process data
fso_births_int_ch <- fso_births_int_ch_raw |>
  # Compute share of Swiss newborns to international mothers
  tidyr::pivot_wider(
    names_from = Beobachtungseinheit,
    values_from = paste0(
      "Szenarien zur Bevölkerungsentwicklung der Kantone 2020-2050",
      " - zukünftige Bevölkerungsentwicklung"
    )
  ) |>
  # use shorter, clearer names
  dplyr::rename(
    # all live births from international mothers
    live_birth_total =
      "Lebendgeburten nach Alter und Staatsangehörigkeit der Mutter",
    # live births of international newborns to international mothers
    live_birth_int = Lebendgeburten
  ) |>
  dplyr::mutate(
    births_int_ch = (live_birth_total - live_birth_int) / live_birth_total
  ) |>
  # Bring variable names and factor levels into the format required later
  dplyr::rename(
    scen = "Szenario-Variante",
    age = Altersklasse,
    year = Jahr
  ) |>
  # change factor levels
  dplyr::mutate(
    scen = dplyr::case_match(
      scen,
      "Referenzszenario AR-00-2020" ~ "reference",
      "'hohes' Szenario BR-00-2020" ~ "high",
      "'tiefes' Szenario CR-00-2020" ~ "low"
    )
  ) |>
  # remove unnecessary variables
  select(year, scen, births_int_ch) |>
  dplyr::arrange(year, scen)
```

## All FSO parameters

Now we can merge "number of people" and "rate" parameters, make the data frame
wider, and compute the required parameter `inter-cantonal net migration`:

```{r all in one, echo = T, eval=FALSE, message=F}
projection_parameters <- dplyr::full_join(fso_rates, fso_numbers) |>
  tidyr::pivot_wider(names_from = fso_parameter, values_from = value) |>
  # compute inter-cantonal net migration
  dplyr::mutate(mig_ch = interc_imm - interc_emi) |>
  left_join(fso_births_int_ch, by = c("year", "scen")) |>
  # add mandatory column spatial_unit
  dplyr::mutate(spatial_unit = "Aargau") |>
  # remove unnecessary variables
  dplyr::select(-c(Kanton, intercant, emi_n, interc_imm, interc_emi)) |>
  dplyr::arrange(year)
```


             
``` {r save-data, message = FALSE, echo=FALSE, eval = FALSE}
# Save data once
save(projection_parameters, file='../vignettes/prepare_projection_parameters.rda')

```             

``` {r load-data, message = FALSE, echo=FALSE, eval = TRUE}
# Load data
load(file='prepare_projection_parameters.rda')
```             

Show parameters for one demographic group for the year 2024:

```{r all in one2, echo = T, eval=TRUE, message=F}
projection_parameters |>
  dplyr::filter(year == 2024 & sex == "f" & nat == "int" & age == 0) |>
  DT::datatable()
```

## Population

In addition to the parameters, the projection function `propop` also requires a 
starting population. To prepare the corresponding query, we again start with the 
metadata:

```{r get meta data population, echo= T, eval=FALSE, results='hide', message=FALSE, warning=FALSE}
metadata_pop <- BFS::bfs_get_metadata(number_bfs = "px-x-0102010000_101")

metadata_pop_tidy <- metadata_pop |>
  select(-valueTexts) |>
  unnest_longer(values) |>
  mutate(
    valueTexts = metadata_pop |>
      select(valueTexts) |>
      unnest_longer(valueTexts) |>
      pull(valueTexts)
  ) |>
  select(code, text, values, valueTexts, everything())

```

We can now specify which levels of the variables we want:

```{r filter dimensions 3, echo = T, eval = FALSE, results='hide', message=FALSE, warning=FALSE}
# Specify the elements to download
dim1 <- metadata_pop_tidy |>
  dplyr::filter(
    text == "Kanton (-) / Bezirk (>>) / Gemeinde (......)" & # Canton
      valueTexts %in% c("- Aargau")
  )

dim2 <- metadata_pop_tidy |>
  dplyr::filter(
    text == "Jahr" & # year
      valueTexts %in% c("2018")
  )

dim3 <- metadata_pop_tidy |>
  dplyr::filter(
    text == "Bevölkerungstyp" & # permanent
      valueTexts %in% "Ständige Wohnbevölkerung"
  )

dim4 <- metadata_pop_tidy |>
  dplyr::filter(
    text == "Staatsangehörigkeit (Kategorie)" & # nationality
      valueTexts %in% c("Schweiz", "Ausland")
  )

dim5 <- metadata_pop_tidy |>
  dplyr::filter(
    text == "Geschlecht" & # sex
      valueTexts %in% c("Mann", "Frau")
  )

dim6 <- metadata_pop_tidy |>
  dplyr::filter(
    text == "Alter" & # age
      !(valueTexts %in% "Alter - Total")
  ) # exclude "Total"

# build dimensions list object
dimensions <- list(
  dim1$values,
  dim2$values,
  dim3$values,
  dim4$values,
  dim5$values,
  dim6$values
)
# add names
names(dimensions) <- c(
  unique(dim1$code),
  unique(dim2$code),
  unique(dim3$code),
  unique(dim4$code),
  unique(dim5$code),
  unique(dim6$code)
)
```

Using the above specifications, we can **download** the FSO "population" 
as follows:

```{r get pop, eval = FALSE, message=FALSE}
# Download population
fso_pop_raw <- BFS::bfs_get_data(
  number_bfs = "px-x-0102010000_101", # reference scenario
  query = dimensions
)
```



            
``` {r save-data2, message = FALSE, echo=FALSE, eval = FALSE}
# Save data once
save(fso_pop_raw, file='../vignettes/prepare_fso_pop_raw.rda')

```             

``` {r load-data2, message = FALSE, echo=FALSE, eval = TRUE}
# Load data
load(file='prepare_fso_pop_raw.rda')
```             

We now process the data to ensure that the population data conforms to 
the structure expected in `propop::propop()`:

```{r prep pop, eval = TRUE, message=FALSE}
# Bring variable names and factor levels into the format required later
starting_population <- fso_pop_raw |>
  dplyr::select(-"Bevölkerungstyp") |>
  dplyr::rename(
    year = Jahr,
    Kanton = "Kanton (-) / Bezirk (>>) / Gemeinde (......)",
    nat = "Staatsangehörigkeit (Kategorie)",
    sex = Geschlecht,
    age = Alter,
    n = "Ständige und nichtständige Wohnbevölkerung"
  ) |>
  # change factor levels
  mutate(
    Kanton = stringr::str_remove_all(Kanton, "- "),
    nat = case_match(
      nat,
      "Schweiz" ~ "ch",
      "Ausland" ~ "int"
    ),
    sex = case_when(
      sex == "Mann" ~ "m",
      sex == "Frau" ~ "f"
    ),
    age = as.numeric(stringr::str_extract(age, "\\d+"))
  ) |>
  dplyr::rename(spatial_unit = Kanton)

starting_population |>
  DT::datatable()
```
\

# Run population projections

Now that the parameters and the starting population are available, we can run 
the population projections (see vignette \`run_projections\` for more details).
The result is shown for one demographic group.
\


```{r project, eval = TRUE, message=FALSE}
# only keep reference scenario
projection_parameters_ref <- projection_parameters |>
  filter(scen == "reference")

# run propop with data from prepare vignette to make sure vignette is okay
results_clean <- propop(
  parameters = projection_parameters_ref,
  year_first = 2019,
  year_last = 2030,
  age_groups = 101,
  fert_first = 16,
  fert_last = 50,
  share_born_female = 100 / 205,
  population = starting_population,
  subregional = FALSE,
  binational = TRUE
)

results_clean |>
  # select demographic group
  dplyr::filter(sex == "f" & nat == "int" & age == 49) |>
  dplyr::mutate(across(n, \(x) sprintf(fmt = "%.0f", x))) |>
  DT::datatable(filter = "top")
```
\
