---
title: "Projections for subregions"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Projections for subregions}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 84
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning=FALSE, message = FALSE, echo = FALSE}
library(propop)
```

```{r setup-display, eval = FALSE}
library(propop)
```


# Overview

This vignette explains how to use `propop::propop()` to perform population 
projections for multiple regions and particularly for subregions within a larger
spatial entity (e.g., municipalities within a canton). 
It discusses two **challenges** that arise when conducting such projections and 
presents potential strategies for addressing them:  

1. How to obtain the **required input data** for spatial entities below 
the level of cantons?
2. How to account for **migration between subregions**? 

For more background and general information about the required input data, see 
[this vignette](project_single_region.html)). 

# Projection parameters for subregions

Numeric information about demographic processes such as births, migration, or 
deaths are an essential prerequisite for population projections. However, in 
Switzerland this information is typically only available for cantons but not 
for spatial entities at smaller scales. 

Supplying input data for spatial units at the sub-cantonal level (e.g. 
municipalities) can be straightforward for data expressed as rates (e.g. 
mortality rate). The simplest approach is to use the same figures for the 
subregions as for the canton (unless the figures are implausible for some
theoretical or empirical reason). The task is more demanding, though, if 
you want to alter rates for subregions or when you need to downscale input data 
expressed as "number of people". While we don't offer a solution to adjust rates
(yet), the next two sections show two possibilities to distribute the cantonal 
"number of people" estimates among subregions.

## Distribution of people according to population size

The simplest approach to allocating canton-wide "number of people" estimates to 
subregions is to determine each subregion's population size relative to the 
canton's total population and distribute the numbers accordingly. To put it 
somewhat simplistically, if a municipality represents 10% of the canton's 
population, it should proportionally receive 10% of the canton's incoming 
immigrants. The approach described below is more sophisticated in that it uses 
the spatial units' shares per demographic group but the core idea is the same.

Let us look at a concrete, numeric example. For the sake of simplicity, we use 
the data included in the package to create small, fictitious input data with 
five regions:

Input parameters (after this step, the rates and numbers for the 
subregions are still identical to those for the whole canton): 

```{r data-parameters} 
# FSO parameters for fictitious subregions
fso_parameters_sub <- fso_parameters |>
  # duplicating rows 5 times
  tidyr::uncount(5) |>
  # create 5 subregions
  dplyr::mutate(spatial_unit = rep(1:5, times = nrow(fso_parameters))) |>
  dplyr::mutate(spatial_unit = as.character(spatial_unit))
```

Population data: 

```{r data-population}
# Generate 5 random "cuts" to distribute the original population; 
# avoid extreme values with a range of 0.1 to 0.5
cut_1 <- {
  set.seed(1)
  round(runif(1, min = 0.1, max = 0.5), digits = 2)
}
cut_2 <- {
  set.seed(2)
  round(runif(1, min = 0.1, max = 0.5), digits = 2)
}
cut_3 <- {
  set.seed(3)
  round(runif(1, min = 0.1, max = 0.5), digits = 2)
}
cut_4 <- {
  set.seed(4)
  round(runif(1, min = 0.1, max = 0.5), digits = 2)
}
# make sure everything adds up to 100%
cut_5 <- 1 - cut_1 - cut_2 - cut_3 - cut_4

# Generate population data for five subregions
df_population <- fso_population |>
  # duplicating rows 5 times
  tidyr::uncount(5) |>
  # create 5 subregions
  dplyr::mutate(
    spatial_unit = as.character(rep(1:5, times = nrow(fso_population)))
  ) |>
  dplyr::mutate(
    # Distribute original population according to "cuts"
    n = dplyr::case_match(
      spatial_unit,
      "1" ~ round(n * cut_1),
      "2" ~ round(n * cut_2),
      "3" ~ round(n * cut_3),
      "4" ~ round(n * cut_4),
      "5" ~ round(n * cut_5),
      .default = NA
    ),
    .keep = "all"
  )
```

To calculate the shares, we count the number of people in each demographic group 
across all spatial units (`sum_n`). Next, the `share` is obtained by 
dividing the spatial unit's `n` by the sum of all people across all spatial units in this
demographic group (`sum_n`). Note that the sum of all five region's shares is 1
(e.g., 0.21 + 0.17 + 0.17 + 0.33 + 0.12 = `r 0.21 + 0.17 + 0.17 + 0.33 + 0.12`).

```{r shares-pop-size}
# Calculate shares
df_population_shares <- df_population |>
  dplyr::mutate(sum_n = sum(n), .by = c(nat, sex, age)) |>
  dplyr::mutate(share = n / sum_n)

# Display table
df_population_shares |>
  dplyr::mutate(share = round(share, 3)) |>
  DT::datatable() |>
  DT::formatStyle(c("share"),
                  backgroundColor = DT::styleRow(c(1:5), "#96D4FF", default = "")
  )
```
\ 

Now all required data are available and we can distribute projection parameters
expressed as "number of people" between subregions. 

Let's do this with immigration from other countries (`imm_int_n`) and immigration 
from other cantons (`imm_nat_n`). We first join the data frame containing the 
projection parameters (`fso_parameters_sub`) and the data frame containing the 
shares (`df_population_shares`). Identifiers for demographic groups must be 
present in both data frames. The actual distribution involves only a single line 
per parameter in which the canton-wide number of immigrants (`imm_int_n` and 
`imm_nat_n`) is multiplied by the share (`share`), which in this approach is
identical for both types of immigration.

```{r distribute-pop-size}
parameters_sub_size <- fso_parameters_sub |>
  dplyr::left_join(
    df_population_shares |>
      dplyr::select("spatial_unit", "nat", "sex", "age", "share"),
    by = c("spatial_unit", "nat", "sex", "age")
  ) |>
  dplyr::mutate(
    # Calculate number of incoming people per demographic group and spatial unit
    imm_int_n_distr = imm_int_n * share,
    imm_nat_n_distr = imm_nat_n * share
  )
```

Let's take a closer look at the result of this, focusing on immigration from 
other countries. In the table below, the blue columns `check` and `difference` 
show that the sum of the distributed parameter (`imm_int_n_distr`) adds up to the 
total number of people (`imm_int_n`; i.e., the original figures provided by 
the FSO for the whole canton). 

```{r distr-pop-size-check}
parameters_sub_size |>
  dplyr::mutate(
    check = round(sum(imm_int_n_distr), 0),
    .by = c(year, nat, sex, age)
  ) |>
  dplyr::filter(sex == "m" & nat == "int") |>
  dplyr::mutate(across(c(
    "int_mothers":"emi_nat", "acq", "share":"imm_nat_n_distr"
  ), \(x) round(x, 3))) |>
  dplyr::mutate(difference = imm_int_n - check) |>
  dplyr::select(-scen) |>
  DT::datatable() |>
  DT::formatStyle(
    "imm_int_n",
    backgroundColor = "#ffcc8f"
  ) |>
  DT::formatStyle(
    c("check", "difference"),
    backgroundColor = "#96D4FF"
  )
```
\
To proceed with the projection, the columns with the distributed immigration 
(`imm_int_n_distr` and `imm_nat_n_distr`) need to become the new 
`imm_int_n` and `imm_nat_n`. Otherwise, `propop()` won't recognize the parameters.  

``` {r cleaning}
parameters_sub_size_clean <- parameters_sub_size |>
  dplyr::mutate(
    imm_int_n = imm_int_n_distr,
    imm_nat_n = imm_nat_n_distr
  ) |>
  dplyr::select(-share, -imm_int_n_distr)
```

Now we can run the projection:

```{r project-1}
propop(
  parameters = parameters_sub_size_clean |> 
    dplyr::filter(scen == "reference"),
  year_first = 2025,
  year_last = 2026,
  age_groups = 101,
  fert_first = 16,
  fert_last = 50,
  share_born_female = 100 / 205,
  population = df_population,
  binational = TRUE,
  subregional = FALSE
)
```


## Distribution of people according to past migration

The second approach to distribute "number of people" estimates among subregions 
uses historical migration records. An advantage of this approach is that it can
differentiate between different types of migration (or any number-of-people 
input) and thereby enables adjusting each parameter independently (rather than 
using the same share for adapting all parameters). This is great, for example, if 
immigration from other cantons differs from immigration from other countries.

To illustrate this approach, we use international immigration as an example and 
create a fictitious data frame with five regions. For pedagogical reasons, we 
modify the data a little bit. Spatial unit 1 won't have any international 
immigration among children aged 0-4 years. Consequently, the value in the column 
`hist_imm_int` contains zeros for these entries. This doesn't reflect realistic 
conditions but helps to illustrate the suggested procedure in cases where zero 
observations exist for demographic groups.

With real data, the same can be achieved by summarizing the 
historical migration data. Since real migration usually varies from 
year to year, we advise calculating an average across several years. 
The fictitious data assumes that we have already performed this step. 

```{r data}
set.seed(145)

# Immigration records
df_hist_imm <- tibble::tibble(
  # five fictitious spatial units
  spatial_unit = rep(c("1", "2", "3", "4", "5"), each = 101 * 4),
  # two nationalities
  nat = rep(rep(c("ch", "int"), each = 2 * 101), times = 5),
  # two levels for sex
  sex = rep(rep(c("m", "f"), each = 101), times = 5 * 2),
  # age groups from 0 to 100
  age = rep(0:100, times = 5 * 4)
) |>
  dplyr::mutate(
    # random numbers between zero and 50 are created to mimic historical
    # immigration records observed in Aargau
    hist_imm_nat = sample(0:50, dplyr::n(), replace = TRUE),
    hist_imm_int = sample(0:50, dplyr::n(), replace = TRUE),
    # for one spatial unit, we artificially assign zeros to all children
    # between 0-4 years
    hist_imm_int = ifelse(spatial_unit == 1 & age %in%
                            c(0:4), 0, hist_imm_int)
  )
```

The resulting data frame includes two columns with (fictitious) immigration 
from abroad (` hist_imm_nat`) and from other cantons (`hist_imm_nat`), aggregated
per demographic group:

```{r data-historical-immigration}
df_hist_imm |>
  DT::datatable()
```

\ 

A challenge is that records often vary considerably between years. If the 
patterns are too uneven, future trends may become erratic, especially in 
groups with few people (e.g., small municipalities or small age groups). 

A first step to mitigate this issue should already have happened at this stage, 
namely considering several years and using the arithmetic mean as an estimate
of each demographic group's past migration.

`propop::calculate_shares()` offers an additional remedy. Instead of only using
shares based on 1-year age groups, it is possible to resort to 5-year and 10-year 
age groups, which further smooths out irregular patterns. 

To use this function, you need to provide a data frame containing a column with
the number of immigrants per demographic group (e.g., `hist_imm_nat` or 
`hist_imm_int` as calculated above). The function assigns each 1-year
age group to a 5-year and a 10-year age group (e.g., 2-year olds become part of 
the age_groups 0-4 and 0-9), sums up all observations within the respective 
5-year and 10-year age group (e.g., 0-4 and 0-9 year old Swiss males). These 
sums (`sum_5`and `sum_10` in the table below) are then divided by five or ten, 
respectively, and proportionally assigned to each member of the age group 
(`prop_5` and `prop_10` in the table below). 

To **distribute the canton-wide numbers** among subregions, 
`propop::calculate_shares()` uses the 1-year age group if its mean share is 
larger than zero. If no immigration was recorded for a particular 1-year age 
group (e.g., the 76-year old Swiss males in spatial unit 1), the mean share of 
the corresponding 5-year age group is used (i.e., 75-79-year old Swiss males 
in the case of 76-year old). If the share across the 5-year age group is also 
zero, the 10-year age group is used (this could be 70-79 year old Swiss males). 
In this case, the `prop_10` share of the respective 10-year age group is used for 
both 5-year age groups within the 10-year age group (e.g., the 70-74 and 75-79 
year olds). The variable `use_age_group` indicates which share is suggested by 
this default algorithm.

The final outcome `share` indicates the proportion of the historical total per
demographic group (`n_sum`) that is to be allocated to the respective
spatial unit. (`share =  n / n_sum`).

```{r result-historical-immigration}
data_distr_hist_int <- df_hist_imm |>
  calculate_shares(col = "hist_imm_int", age_group = "age_group_5")

data_distr_hist_int |>
  # Display rounded numbers to save space
  dplyr::mutate(share = round(share, 3)) |>
  DT::datatable() |>
  DT::formatStyle(
    "n",
    backgroundColor = "#96D4FF"
  ) |>
  DT::formatStyle(
    "n_sum",
    backgroundColor = "#96D4FF"
  ) |>
  DT::formatStyle(
    "share",
    backgroundColor = "#007AB8"
  )
```

\ 

Note that summarizing the shares of a demographic group across all spatial units 
always adds up to 1:


```{r share-check}
data_distr_hist_int |>
  dplyr::summarise(sum_share = round(
    sum(share, na.rm = TRUE),
    digits = 1), .by = c(nat, sex, age)) |>
  DT::datatable()
```

\
From now on, the procedure is identical to the first approach that distributed
the number of people according to population size. That is, the share is 
multiplied with the numbers that the FSO estimated for the whole canton.

```{r parameters-distributed}
# In addition to international immigration shares and numbers
# we also need the shares and numbers for national immigration
data_distr_hist_nat <- df_hist_imm |>
  calculate_shares(col = "hist_imm_nat") |>
  # Use unambiguous name
  dplyr::rename(share_imm_nat = share) |>
  # Drop unnecessary variables
  dplyr::select(-c(
    "hist_imm_nat", "hist_imm_int", "age_group_5", "sum_5", "prop_5", 
    "age_group_10", "sum_10", "prop_10", "use_age_group", "n", "n_sum"
  ))

# Join both data frames holding shares
data_distr_hist <- data_distr_hist_int |>
  # Use unambiguous name
  dplyr::rename(share_imm_int = share) |>
  # Drop unnecessary variables
  dplyr::select(-c(
    "hist_imm_nat", "hist_imm_int", "age_group_5", "age_group_10", "sum_10",
    "prop_10", "use_age_group", "n", "n_sum"
  )) |>
  dplyr::left_join(data_distr_hist_nat,
                   by = c("spatial_unit", "nat", "sex", "age")
  )

# Add shares to the data frame that holds the projection parameters
fso_parameters_sub_distr_hist <- fso_parameters_sub |>
  dplyr::left_join(
    data_distr_hist |>
      dplyr::select(
        "spatial_unit", "nat", "sex", "age", "share_imm_int", "share_imm_nat"
      ),
    by = c("spatial_unit", "nat", "sex", "age"),
    relationship = "many-to-one"
  ) |>
  # Compute `n` for subregions, assign values directly to imm_int_n and imm_nat_n
  dplyr::mutate(
    imm_int_n = imm_int_n * share_imm_int,
    imm_nat_n = imm_nat_n * share_imm_nat
  ) |>
  # Remove unnecessary variables
  dplyr::select(-c("share_imm_int", "share_imm_nat"))

# Show result
fso_parameters_sub_distr_hist |>
  # Remove variables for better overview
  dplyr::select(-fso_projection_n, -scen) |>
  head(100) |>
  dplyr::mutate(across(
    c("int_mothers":"imm_nat_n"), \(x) sprintf(fmt = "%.3f", x)
  )) |>
  DT::datatable()
```
Now everything is ready to run the projection: 

```{r project-2}
propop(
  parameters = fso_parameters_sub_distr_hist|> 
    dplyr::filter(scen == "reference"),
  year_first = 2025,
  year_last = 2026,
  age_groups = 101,
  fert_first = 16,
  fert_last = 50,
  share_born_female = 100 / 205,
  population = df_population,
  binational = TRUE,
  subregional = FALSE
)
```

# Migration between subregions

`propop` offers the possibility to account for migration between subregions. To
adjust the population size in each subregion according to past migration,
the column `mig_sub` (= migration in subregions) is required in the parameter 
data frame.

Here we add `mig_sub` as a *fictitious* parameter the parameter input file. (In 
real life you use population register records instead). 

```{r, eval = TRUE}
parameters_sub_mig <- fso_parameters_sub_distr_hist |>
  # Create fictitious migration parameters
  dplyr::mutate(
    mig_sub = dplyr::case_when(
      # Four regions with emigration, 1 region with immigration
      spatial_unit == 1 ~ {
        set.seed(1)
        round(rnorm(1, mean = 0, sd = 0.2), digits = 4)
      },
      spatial_unit == 2 ~ {
        set.seed(2)
        round(rnorm(1, mean = 0, sd = 0.2), digits = 4)
      },
      spatial_unit == 3 ~ {
        set.seed(25)
        round(rnorm(1, mean = 0, sd = 0.2), digits = 4)
      },
      spatial_unit == 4 ~ {
        set.seed(12)
        round(rnorm(1, mean = 0, sd = 0.2), digits = 4)
      },
      TRUE ~ NA
    )
  ) |>
  dplyr::mutate(
    mig_sub = dplyr::case_when(
      spatial_unit == 5 ~ 0 - sum(mig_sub, na.rm = TRUE), TRUE ~ mig_sub
    ),
    # check = sum(mig_sub, na.rm = TRUE),
    .by = c("nat", "sex", "age", "year", "scen")
  ) |>
  dplyr::select(
    nat, sex, age, year, scen, spatial_unit, birthrate, int_mothers, mor,
    emi_int, emi_nat, imm_int_n, imm_nat_n, acq, emi_nat_n, mig_nat_n, mig_sub
  )
```

Now that all required input files are available, we can set `subregional` to
`TRUE` and use `propop::propop()`:

```{r project-3, eval = TRUE}
propop(
  parameters = parameters_sub_mig |> 
    dplyr::filter(scen == "reference"),
  year_first = 2025,
  year_last = 2026,
  age_groups = 101,
  fert_first = 16,
  fert_last = 50,
  share_born_female = 100 / 205,
  population = df_population,
  binational = TRUE,
  subregional = TRUE
)
```
